---
title: <center>Simulating Surface Energy Budget and Evapotranspiration with RMEP</center>
author:
- Yong Yang^[goku@hust.edu.cn],Huaiwei Sun^[hsun@hust.edu.cn]
date: " version 2.0, 11-Feb-2021"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{RMEP-vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Setting up the environment

Loading the *RMEP* library, which contains data and functions.
```{r message=FALSE}
library(RMEP)
```

- The code below show the simple example in the [RMEP](https://github.com/Yangyonghust/RMEP) package. The evapotranspiration(ETMEP, mm/day), sensible heat flux(HMEP, W/m^2^), latent heat flux(EMEP, W/m^2^), ground heat flux(GMEP, W/m^2^) and effective numbers(ef_id) are given in the outputs list.

# Usage
RMEP(Rn,RnL,qs,Ts,type,I,z,...)

# Example

## Calculate evapotranspiration(ET) and heat fluxes
Three variables are required at least in theory for the maximum entropy production (MEP) model inputs, including the net radiation (Rn), specific humidity (qs) and surface temperature (Ts). Net long wave radiation (RnL) variable is also introducing in the RMEP for the case of Water-snow-ice surface in the MEP formulations.
```{r}
RMEP(Rn=300,RnL=100,qs=0.004,Ts=25,I=600,z=2.5,type=1)
```
-  After calculating, the actual ET is 2.546416 (mm/day), the H, E and G are 136.7877, 72.20737, 91.00497 (W/m^2^) respectively, and the effective number is 1. Parameter "Type" represents the condition of surface land cover, type 1 for bare soil surface or short canopy, 2 for dense canopy and 3 for Water-snow-ice surface, and the default is 1.Parameter "I" (constant) is the media thermal inertial (units:TIU),default is 600 TIU. Parameter "z" (constant) is the theoretical height above surface (unit:m), and default is 2.5 m.

-  If you want to calculate potential evapotranspiration (PET), only the Rn and Ts two variables are needed.
```{r}
RMEPPET(Rn=300,RnL=100,Ts=25,type=2)

```
-  The PET is 7.386154 mm/day.

## Calculate the specific humidity for MEP model input

To calculate specific humidity(qs), three variables are needed, including: Air temperature(TA, unit: deg C), Relative humidity(RH, unit: percent), and Atmospheric pressure(PA, unit: kPa).
```{r}
Shum(TA=20,PA=101,RH=50)

```
- The qs is 0.007311296 kg/kg.

## Calculate the saturated specific humidity for MEP model input
Only one variable: surface temperature(Ts, unit: deg C) are required for calculating saturated specific humidity.
```{r}
SShum(Ts=20)

```
- The saturated specific humidity is 0.01464032 kg/kg.

## Analysis of time series data
A time series based dataset "datademo" is used for analysis with RMEP, which contains 8 variables during 1-30, June 2016 with half-hourly time step. 
```{r fig.height=4, fig.width=7, warning=FALSE}
#Load example data
data("datademo") 
Data <- datademo
Data[1:10,] #View the data and variables
qs <- Shum(TA <- Data$TA,PA <- Data$PA,RH <- Data$RH) #Calculate specific humidity
Rn <- Data$NETRAD;RnL <- Data$NETRAD #RnL is not used in type=1 so let it equals to Rn
Ts <- Data$TS #Surface temperature
output <- RMEP(Rn,RnL,qs,Ts,type=1) #The results is obtained in the "output" list
simH <- t(output$HMEP);simE <- t(output$EMEP);simG <- t(output$GMEP)#The results of H, E and G as a vector

#Package "ggplot2" is used to produce time series plot between RMEP results and observations
library(ggplot2)
lm(simH~Data$H) #Linear fit between calculated and observed H
qplot(Data$H,simH)+geom_point(col="blue")+geom_smooth(method="lm",formula = y~x,col="red",lty=1)+
  geom_abline(intercept=0, slope=1,col="black",lty=5,lwd=1)+labs(x="Hobs",y="HMEP",title = "Ro4")+
  annotate("text",x=200,y=50,label="y=0.93x+23.29")  #Scatter plot of MEP estimated and observed H

# Time series change between RMEP calculations and measurements
xlab <- seq(1,length(Data$H),by=1)
timeseries <- data.frame(xlab,Data$H,simH)
fill=c("black","red")
Hplot <- ggplot(timeseries,aes(x=xalab))+geom_line(aes(x=xlab,y=Data$H,color="Measured"))+geom_line(aes(x=xlab,y=simH,color="RMEP"))+labs(x="",y="H(W/m2)",title="Ro4")
Hplot # Time series of calculated H and measured H

Eplot <- ggplot(timeseries,aes(x=xalab))+geom_line(aes(x=xlab,y=Data$LE,color="Measured"))+geom_line(aes(x=xlab,y=simE,color="RMEP"))+labs(x="",y="E(W/m2)",title="Ro4")
Eplot # Time series of calculated E and measured E

Gplot <- ggplot(timeseries,aes(x=xalab))+geom_line(aes(x=xlab,y=Data$G,color="Measured"))+geom_line(aes(x=xlab,y=simE,color="RMEP"))+labs(x="",y="G(W/m2)",title="Ro4")
Gplot # Time series of calculated G and measured G

```

# Software details
This tutorial was built under: 

```{r echo=FALSE}
sessionInfo()$platform
sessionInfo()$R.version$version.string 
paste("RMEP", sessionInfo()$otherPkgs$RMEP$Version)
```

# Version history

* v0.2: Feb 2021
* v0.1: Dec 2020

